The importance of measurement uncertainty in ecological management

### 2018-08-01 manuscripts/manuscript.Rmd
fix eqn, sp
revisions...

$$\begin{equation}
y_t = \xi_t x_t 
\end{equation}$$

@@

$$\begin{equation}
\max_{ \lbrace h_t \rbrace } \mathbb{E} \left \{ \sum_{t=0}^{\infty} \gamma^t \cdot h_t \right \}
\end{equation}$$

@@

$$\begin{equation}
x_{t+1} = \varepsilon_t (x_t-h_t) r  \left(1 - \frac{(x_t-h_t)}{K}\right) 
\end{equation}$$

@@

MSE

@@

So far, attempts to provide a more formal basis for managing uncertainty than the heuristic adjustment of catch limits described above have largely foundered in a series of paradoxes.  Without uncertainty, the theoretical policies are quite intuitive: The result derived by @Clark1973 for an optimal dynamic strategy to replace the equilibrium solution of MSY can be summarized as: If the stock is most productive at $B_{MSY}$, then obtain $B_{MSY}$ as quickly as possible. Thus, at any level below $B_{MSY}$, the economically optimal thing to do is to completely shut down the fishery, while above this stock, harvests greater than $H_{MSY}$ will be needed to bring the stock back to $B_{MSY}$.  (This intuition must be adjusted slightly in the case of economic discounting of future profits, but is otherwise quite general, see @Clark1990).  This strategy is known as *constant escapement* (CE), since a constant sock size $B_{MSY}$ escapes harvest each year.  Because CE and MSY converge to the *same long-term biomass and same long-term average yield*, CE can appear to be merely a more formal justification for MSY.  However, in practice the two approaches are very different: becuase natural populations fluctuate year-to-year, under CE, a different harvest quota must be set each year to achieve the constant escapement.  While theoretical results have proven that CE and not MSY is optimal for fluctuating populations [@Reed1979], real-world examples of CE-regulated fisheries are limited primarily to salmon.  As we shall see, despite being insensitive to natural stochasticity, CE is very sensitive to measurement error -- thus it is perhaps not surprising that CE is not found in other marine fisheries where population sizes may be harder to estimate accurately.

Intuitively, any uncertainty, either in the population growth model (stochasticity), or measurement error, should seem to justify harvesting fewer fish, as is built into the TAC limits discussed above. Instead, existing theory has presented only the paradoxical results that uncertainty either shoud not matter at all, or that we should increase rather than decrease harvests in response to increased uncertainty.

@@

The first of these we shall refer to as Reed's Paradox for stochastic growth, owing to the mathematical proof provided in @Reed1979 which demonstrates that, under sufficiently general assumptions, the optimal escapement $S$ for a population under stochastic growth, is identical to the optimal escapement $D$ of a deterministic population, $S = D = B_{MSY}$.  This surprising result suggests that in going from a world where a manager has perfect knowledge of absolutely everything into a scenario where the manager faces considerable uncertainty about the future state of the world, no additional precaution is needed.  This also provides the manager with a remarkably convenient mechanism for determining the optimal harvest policy: rather than rely on computationally intensive Stochastic Dynamic Programming to determine the optimal policy, a simple derivative will suffice, which is all that is needed to solve the maximum of the deterministic problem.

@@

Despite these notes of caution, both @Clark1986 and @Sethi2005 ultimately attempt to rationalize this counter-intuitive conclusion rather than reject it.  Others have also attempted to introduce measurement error, but always relying on various approximations that similarly alter the problem. The most common such assumption is to assume from the outset that the solution must be of 'constant-escapement' type [@Ludwig1981; @Roughgarden1996; @Engen1997;  @Moxnes2003].  Here we will illustrate the use of powerful modern algorithms that will permit a more direct solution to the case of measurement uncertainty.  In addition to resolving this paradox, we will demonstrate this direct solution also out-performs current more heuristic rules of thumb such as those currently used in fisheries management to account for uncertainty.

@@

Why does the introduction of measurement uncertainty make the decision problem so much harder?  Consider the thought experiment of managing a fish stock where in each of the previous four years stock assessments have put the population size at, say, 51, 54, 49, 51, (say, measured in percent of carrying capacity, $K$). Under a constant escapement target of 50, the harvest quota would be 1, 4, 0, 1, respectively.  Note that this control rule depends only on the measurement for the year in question, which is assumed to be measured without error. If we then measure a stock size of 75, an assumption that this measurement is made without error would justify the sudden increase to a quota of 25. Yet if we admit the possibility of measurement error, this sudden bump starts to look very suspicious in light of all the previous observations.  Perfect measurements let us exploit the Markov property of the model -- the most recent measurement tells us all we need to know.  In contrast, the introduction of measurement uncertainty breaks this Makovian assumption: a rational decision maker would suspect this 75 to be an over-estimate based on prior observations.

As discussed above, absent the short-cut provided by Reed's theorems -- solving for the optimal management policy in a stochastically fluctuating population -- involves a Markov Decision Process (MDP), commonly referred to in ecological literature by its solution method, Stochastic Dynamic Programming (SDP) [@Marescot2013]. The constant escapement (CE) strategy considered here is the result of such an MDP or SDP analysis.  Once measurement uncertainty is introduced, the problem becomes a Partially Observed Markov Decision Process (POMDP, not be confused with a partially observed or hidden Markov model, HMM, in which no decision process is involved).

Precise definitions of both MDP and POMDP problems can be found in @Williams2011, or in classic work from the engineering and robotics community [e.g. @Smallwood1973 and @Sondik1978].

The POMDP problem is posed identically to that of the MDP problem, with the addition of an observation process  The POMDP problem for fisheries question considered here can be summarized as follows:

@@

The optimization problem is to select the action $a_t$ that will maximize the net present utility over all time. Future utility may be discounted by a factor $\gamma$, so that a value  $V$ in $t$ year is valued at $\gamma^t V$ today.  Numerically, each of these functions are defined over a discrete set of possible states, observations, and actions, and can thus be represented as a collection of matrices or tensors.  While the solution method for MDP problems, stochastic dynamic programming (SDP) is well known and readily implemented in ecological problems [e.g @Mangel1988; @Marescot2013], the POMDP problem cannot be efficiently solved using SDP. While not unknown to the conservation literature [@Williams2011], algorithms for POMDP have historically scaled quite poorly, and their application has been restricted to contexts with only a handful of possible states and actions [e.g. @Chades2008; @Chades2011; @Fackler2014; @Fackler2014b].  POMDP algorithms have a substantial literature starting almost contemporaneously to the work of @Reed1979 (see foundational papers by @Smallwood1973 and @Sondik1978) and remain an active area in artificial intelligence [e.g. @Kaelbling1998; @Pineau2003]. By adapting more recent algorithmic developments [@Kurniawati2008] from that field we are able to find solutions for the considerably more complex problems such as the case of fisheries management, which requires on the order of 100 states and actions to provide sufficient numerical resolution. As the details for solving such problems numerically are already well documented we will not rehash them here, but instead provide a preliminary R package [@sarsop-pkg] which provides an implementation of the SARSOP algorithm and the fisheries POMDP problem, which can be used to replicate any of the results presented here and explore other variations.  Details of the analysis presented here including annotated code to reproduce and further explore all of the following results is provided in the supplementary material.

@@

Differences in attention to prior beliefs,determined by prior observations, drive the differences between the POMDP strategy and the other strategies and can resolve the uncertainty paradox.  While the catch quota under both TAC and CE strategies can be completely determined given the most recent observation of the stock size by using the policy curves shown in Figure \ref{policy}, this is not the case for the POMDP approach. This fundamental difference is key to understanding the difference in performance and resolving the paradox of uncertainty.  The POMDP policy cannot be specified by the most recent observation alone. Instead, the POMDP policy depends on all prior observations, not just the most recent.  The reason for this complexity comes from the Markov property.  Observations of the state in the perfectly observed system satisfy the Markov property: once we have measured the current biomass exactly, we cannot get any better estimate of the current stock size by studying older measurements. When measurements are uncertain this is no longer the case: intuitively, by comparing the most recent measurement to previous observations we may be able to infer when any given measurement is unusually high or unusually low.  POMDP formalizes this intuition by capturing the information from all previous observations into a *prior belief*.  This prior belief is updated after every subsequent action and observation in accordance with Bayes Law.  The mechanics of this process are well documented in the extensive literature on POMDPs [e.g. @Smallwood1973; @Sondik1978; @Kurniawati2008; @Williams2011], but for our purposes it is sufficient to observe how this evolving prior belief serves to continually adjust the POMDP policy.  Supplemental Figure S9 (Appendix A) illustrates how the prior belief evolves in response to subsequent observations and actions over the course of an individual simulation.

>>>>

\begin{equation}
y_t = \xi_t x_t 
\end{equation}

@@

\begin{equation}
\max_{ \lbrace h_t \rbrace } \mathbb{E} \left \{ \sum_{t=0}^{\infty} \gamma^t \cdot h_t \right \}
\end{equation}

@@

\begin{equation}
x_{t+1} = \varepsilon_t (x_t-h_t) r  \left(1 - \frac{(x_t-h_t)}{K}\right) \label{Gordon-Schaefer}
\end{equation}

@@

<!--- FIXME MSE discussion HERE!! -->

@@

So far, attempts to provide a more formal basis for managing uncertainty than the heuristic adjustment of catch limits described above have largely foundered in a series of paradoxes.  Without uncertainty, the theoretical policies are quite intuitive: The result derived by @Clark1973 for an optimal dynamic strategy to replace the equilibrium solution of MSY can be summarized as: If the stock is most productive at $B_{MSY}$, then obtain $B_{MSY}$ as quickly as possible. Thus, at any level below $B_{MSY}$, the economically optimal thing to do is to completely shut down the fishery, while above this stock, harvests greater than $H_{MSY}$ will be needed to bring the stock back to $B_{MSY}$.  (This intuition must be adjusted slightly in the case of economic discounting of future profits, but is otherwise quite general, see @Clark1990).  This strategy is known as *constant escapement* (CE), since a constant sock size $B_{MSY}$ escapes harvest each year.  Because CE and MSY converge to the *same long-term biomass and same long-term average yield*, CE can appear to be merely a more formal justification for MSY.  However, in practice the two approaches are very different: becuase natural populations fluctuate year-to-year, under CE, a different harvest quota must be set each year to achieve the constant escapement.

In the face of a naturally fluctuating population, previous arguments made by @Gordon1954; @Schaefer1954 or @Clark1973 no longer hold.  The optimal harvest of a population under a model such as the stochastic Gordon Schaefer, Eq \eqref{Gordon-Schaefer}, is an example of a Markov Decision Problem (MDP). This is not to be confused with the the problem of *estimating* parameters or posteriors of a Markov process model, e.g. through typical Bayesian analysis -- recall this is a decision problem, in which the model has already been estimated and now we are tasked with choosing a sequence of future harvests based on that model.  Such MDP problems are commonly referred to in ecological literature by its solution method, Stochastic Dynamic Programming (SDP) [@Marescot2013], widely used in both ecological management and animal behavior [@Mangel1985; @Mangel1988].  However, thanks to a mathematical proof provided by @Reed1979, fisheries managers have largely been able to bi-pass the need for SDP.  But this convenient result also opened the door to paradox of uncertainty that would persist for the next four decades.

@@

 @Reed1979 proved that, under sufficiently general assumptions, the optimal escapement $S$ for a population under stochastic growth, is identical to @Clark1973's optimal escapement $D$ for a deterministic population, $S = D = B_{MSY}$.  This surprising result suggests that in going from a world where a manager has perfect knowledge of absolutely everything into a scenario where the manager faces considerable uncertainty about the future state of the world, no additional precaution is needed.

@@

Despite these notes of caution, both @Clark1986 and @Sethi2005 ultimately attempt to rationalize this counter-intuitive conclusion rather than reject it.  Intuitively, uncertainty in the population growth model (stochasticity) or measurement error, would seem to justify harvesting fewer fish. Instead, existing optimal control theory has presented only the paradoxical results that uncertainty either shoud not matter at all, or that we should increase rather than decrease harvests in response to increased uncertainty. Numerous other attempts have been made to resolve this paradox through alternative assumptions,tThe most common being to assume from the outset to look only for 'constant-escapement' type polcies [@Ludwig1981; @Roughgarden1996; @Engen1997;  @Moxnes2003]. Owing to the "mathematical difficulty" Clark first observed, none have attempted a direct solution which we will present here.

In practice, TAC-managed fisheries are more common that CE-managed fisheries, at least in the United States. CE-based regulation is limited primarily to salmon, species for which population estimates may be more precise and thus closer to Reed's assumption.  We are not aware of any fishery that has put Clark's Paradox into practice, harvesting more in response to increased uncertainty.  As we shall illustrate, this is particularly fortunate; though approaches that properly account for both forms of uncertainty can perform even better.

@@

Why does the introduction of measurement uncertainty make the decision problem so much harder?  Consider the thought experiment of managing a fish stock where in each of the previous four years stock assessments have put the population size at, say, 51, 54, 49, 51, (say, measured in percent of carrying capacity, $K$). Under a constant escapement target of 50, the harvest quota would be 1, 4, 0, 1, respectively.  Note that this control rule depends only on the measurement for the year in question, which is assumed to be measured without error. If we then measure a stock size of 75, an assumption that this measurement is made without error would justify the sudden increase to a quota of 25. Yet if we admit the possibility of measurement error, this sudden bump starts to look very suspicious in light of all the previous observations.  Perfect measurements let us exploit the Markov property of the model -- the most recent measurement tells us all we need to know.  In contrast, the introduction of measurement uncertainty breaks this Markovian assumption: a rational decision maker would suspect this 75 to be an over-estimate based on prior observations. Under imperfect observations, we no longer have a Markov Decision Process (MDP) over the observed state space.  Rather, we have a Partially Observed Markov Decision Process (POMDP). A POMDP policy cannot be expressed as a policy function at all (much less as policy function with a single constant-escapement target, as much previous work has assumed).  Instead, a POMDP policy can be thought of as depending not only on the most recent observation, but the recent observation combined with a *prior belief* about what state the system is in, which itself is determined by all past observations. In this manner, the POMDP can consider how much weight to give the most recent observation based on how it compares to prior beliefs.

Algorithmic innovations from the robotics community have now made it possible to numerically solve non-trivial POMDP problems. The need to consider all past observations becomes particularly demanding given the forward-looking nature of a decision problem.  Not only must we consider all possible sequences of future actions over all possible sequences of future states, but we must add to that all possible sequences of future measurements. The computational complexity involved has historically limited POMDP applications to contexts with only a handful of possible states and actions [e.g. @Chades2008; @Chades2011; @Fackler2014; @Fackler2014b], insufficient to capture fisheries models appropriately.  Meanwhile, methods to solve decision problems with imperfect information have advanced steadily in the field of robotics and artificial intellengence.  Almost contemporaneous with Reed's work,  papers by @Smallwood1973 and @Sondik1978 laid mathematical foundations for algorithms that could efficiently solve POMDP problems [e.g. @Kaelbling1998; @Pineau2003].  More recently, driven by demands in areas such as autonomous vehicle decision making, newer point-based approximation algorithms such as SARSOP [@Kurniawati2008] have made it possible to solve POMDPs with 100s of states and actions.  We have adapt this algorithm for application to the fisheries context, and provide an efficient implementation as a preliminary R package [@sarsop-pkg] which can be used to replicate and further explore results presented here. Details of the analysis presented here including annotated code to reproduce and further explore all of the following results is provided in the supplementary material.

Precise definitions of both the (fully observed) MDPs and POMDPs (not to be confused with Markov Process Models and Hidden Markov Process Models -- methods for which estimate the model but do not involve the feedback of any decision process) in an ecological context can be found in @Williams2011, or in classic work from the engineering and robotics community [e.g. @Smallwood1973 and @Sondik1978].  For convenience, we summarize the POMDP definition and common notation here.  The POMDP problem is posed identically to that of the MDP problem, with the addition of an observation process  The POMDP problem for fisheries question considered here can be summarized as follows:

@@

The optimization problem is to select the action $a_t$ that will maximize the net present utility over all time. Future utility may be discounted by a factor $\gamma$, so that a value  $V$ in $t$ year is valued at $\gamma^t V$ today.  Numerically, each of these functions are defined over a discrete set of possible states, observations, and actions, and can thus be represented as a collection of matrices or tensors.

@@

Differences in attention to prior beliefs, determined by prior observations, drive the differences between the POMDP strategy and the other strategies and can resolve the uncertainty paradox.  While the catch quota under both TAC and CE strategies can be completely determined given the most recent observation of the stock size by using the policy curves shown in Figure \ref{policy}, this is not the case for the POMDP approach. This fundamental difference is key to understanding the difference in performance and resolving the paradox of uncertainty.  The POMDP policy cannot be specified by the most recent observation alone. Instead, the POMDP policy depends on all prior observations, not just the most recent.  The reason for this complexity comes from the Markov property.  Observations of the state in the perfectly observed system satisfy the Markov property: once we have measured the current biomass exactly, we cannot get any better estimate of the current stock size by studying older measurements. When measurements are uncertain this is no longer the case: intuitively, by comparing the most recent measurement to previous observations we may be able to infer when any given measurement is unusually high or unusually low.  POMDP formalizes this intuition by capturing the information from all previous observations into a *prior belief*.  This prior belief is updated after every subsequent action and observation in accordance with Bayes Law.  The mechanics of this process are well documented in the extensive literature on POMDPs [e.g. @Smallwood1973; @Sondik1978; @Kurniawati2008; @Williams2011], but for our purposes it is sufficient to observe how this evolving prior belief serves to continually adjust the POMDP policy.  Supplemental Figure S9 (Appendix A) illustrates how the prior belief evolves in response to subsequent observations and actions over the course of an individual simulation.
